---
title: "Activity: Graphing Regions on x-y Plane"
author: "Jibo Shen"
output: 
  github_document:
    toc: false
---

```{r setup, include=FALSE}
# This chunk sets up the environment.
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.align="center")

library(ggplot2)
library(gridExtra)

# Common plotting theme for consistency
common_theme <- theme_minimal() + 
                theme(axis.title = element_text(face="bold"),
                      plot.title = element_text(face="bold", hjust = 0.5))

```

Visualizing the region where a function is defined is a critical skill. This activity reviews how to sketch common regions defined by inequalities on the $x$-$y$ plane.

---

### Introduction to the $x$-$y$ Plane (The Cartesian Plane)

The $x$-$y$ plane is a two-dimensional flat surface formed by the intersection of two perpendicular number lines. It allows us to pinpoint the exact location of any point or shape using a pair of numbers. 

Here are the key components:

* **The $x$-axis:** The horizontal number line. Moving to the right from the center gives positive values, and moving to the left gives negative values.
* **The $y$-axis:** The vertical number line. Moving up from the center gives positive values, and moving down gives negative values.
* **The Origin:** The exact center of the plane where the $x$-axis and $y$-axis intersect. It represents the starting point for measuring and has the coordinates $(0,0)$.
* **Coordinates $(x, y)$:** Every point on the plane is defined by an ordered pair. The first number ($x$) tells you how far to move horizontally along the $x$-axis, and the second number ($y$) tells you how far to move vertically parallel to the $y$-axis.
* **The Four Quadrants:** The two intersecting axes divide the plane into four distinct sections, called quadrants. They are numbered counterclockwise using Roman numerals, starting from the top right:
    * Quadrant I: Both $x$ and $y$ are positive ($x > 0, y > 0$).
    * Quadrant II: $x$ is negative, $y$ is positive ($x < 0, y > 0$).
    * Quadrant III: Both $x$ and $y$ are negative ($x < 0, y < 0$).
    * Quadrant IV: $x$ is positive, $y$ is negative ($x > 0, y < 0$).

```{r}
# --- 0. The x-y Plane Introduction ---
ggplot() +
  # Draw axes
  geom_hline(yintercept = 0, size = 1) +
  geom_vline(xintercept = 0, size = 1) +
  # Mark the Origin
  geom_point(aes(x = 0, y = 0), size = 3, color = "red") +
  annotate("text", x = 0.3, y = -0.3, label = "Origin (0,0)", color = "red", fontface = "bold") +
  # Label Axes
  annotate("text", x = 4.5, y = 0.3, label = "x-axis", fontface = "bold") +
  annotate("text", x = 0.4, y = 4.5, label = "y-axis", fontface = "bold") +
  # Label Quadrants
  annotate("text", x = 2.5, y = 2.5, label = "Quadrant I\n(+, +)", size = 5, color = "blue") +
  annotate("text", x = -2.5, y = 2.5, label = "Quadrant II\n(-, +)", size = 5, color = "blue") +
  annotate("text", x = -2.5, y = -2.5, label = "Quadrant III\n(-, -)", size = 5, color = "blue") +
  annotate("text", x = 2.5, y = -2.5, label = "Quadrant IV\n(+, -)", size = 5, color = "blue") +
  xlim(-5, 5) + ylim(-5, 5) +
  labs(title = "The x-y Plane (Cartesian Coordinate System)") +
  theme_minimal() +
  theme(axis.title = element_blank())
```

### What is a Region?

On the $x$-$y$ plane, an equation with an equals sign (like $y = 2x$ or $x^2 + y^2 = 4$) describes a specific boundary—a 1D line or curve. 

A **region** is a two-dimensional area on the plane. It consists of a collection of points $(x,y)$ that satisfy an **inequality** (like $<$, $>$, $\le$, or $\ge$). For example, $y \le x$ is the entire area of the plane that falls exactly on or anywhere below that line.

When you are asked to graph a region, you are essentially looking for the overlapping area where all the given conditions (inequalities) are true at the same time, and then you shade that area in.

---

### Rectangular Regions
Inequalities: $a \le x \le b$ and $c \le y \le d$

**How to Sketch:**

1.  Vertical Bounds: Draw vertical lines passing through $x = a$ and $x = b$.
2.  Horizontal Bounds: Draw horizontal lines passing through $y = c$ and $y = d$.
3.  Region: The rectangle is the overlapping area trapped between these four lines.

Example:
Sketch the region where $1 \le x \le 3$ and $1 \le y \le 2$.

* Vertices: $(1,1), (3,1), (3,2), (1,2)$.

```{r}
library(ggplot2)

# --- 1. Rectangular Region ---
rect_data <- data.frame(xmin=1, xmax=3, ymin=1, ymax=2)
ggplot() +
  geom_rect(data=rect_data, aes(xmin=xmin, xmax=xmax, ymin=ymin, ymax=ymax), 
            fill="lightblue", color="black", alpha=0.5) +
  xlim(0, 4) + ylim(0, 3) +
  labs(title = "Rectangular Region", x = "x", y = "y") +
  theme_minimal()
```

---

### Triangles (Bounded by Axes)
Inequalities: $Ax + By \le C$ (with $x \ge 0, y \ge 0$)

**Key Concept:**

The equation $Ax + By = C$ represents a straight line. This line acts as a boundary that **splits the entire $x$-$y$ plane into two separate regions**:
1.  One side where $Ax + By < C$.
2.  The other side where $Ax + By > C$.
To graph the inequality, you simply need to draw the line and determine which "half" of the plane you are in.

**How to Sketch:**

1.  Find Intercepts: Treat the inequality as a line equation ($Ax + By = C$).
    * Set $y = 0$ to find the $x$-intercept.
    * Set $x = 0$ to find the $y$-intercept.
2.  Draw the Line: Plot these two intercepts on the axes and connect them with a straight line.
3.  Determine the Region: An easy way to do so is picking a point on the plane to test.
    * Pick a random point **not** on the line to test. $(0,0)$ is usually the easiest choice.
    * Plug $(0,0)$ into the inequality: Is $A(0) + B(0) < C$?
    * If **YES** ($0 < C$), then the region *contains* the origin (shade the side with $(0,0)$).
    * If **NO**, then the region is on the *other* side of the line.
    * If $(0,0)$ is on the line, simply pick another point that is easy to evaluate.

Example:
Sketch the region $2x + 3y \le 6$ in the first quadrant ($x,y \ge 0$).

* Intercepts:
    * $x$-intercept: $2x=6 \implies x=3$. Point $(3,0)$.
    * $y$-intercept: $3y=6 \implies y=2$. Point $(0,2)$.
* Test Point $(0,0)$:
    * Check: $2(0) + 3(0) = 0$.
    * Is $0 < 6$? **Yes.**
    * Therefore, we shade the region *towards* the origin.

```{r}
# --- 2. Triangle (One Line) ---
tri_data <- data.frame(x = c(0, 3, 0), y = c(0, 0, 2))
ggplot(tri_data, aes(x, y)) +
  geom_polygon(fill="lightgreen", color="black", alpha=0.5) +
  xlim(0, 4) + ylim(0, 3) +
  labs(title = "Triangular Region (2x + 3y <= 6)", x = "x", y = "y") +
  theme_minimal()
```


---

### Triangles (Bounded by Two Lines + Axis)
Inequalities: $y \le L_1(x)$, $y \le L_2(x)$, and $y \ge 0$.

**How to Sketch:**

1.  Find the Peak (Intersection): Set the two line equations equal to each other ($L_1(x) = L_2(x)$) to find where they intersect.
2.  Find the Base: Find where each line crosses the horizontal axis.
3.  Region: Connect the intersection point to the two base points.

Example:
Sketch the region bounded by $y = x$, $y = 2 - x$, and $y = 0$.

* Intersection: $x = 2 - x \implies 2x = 2 \implies x = 1$. Peak at $(1, 1)$.
* Base:
    * $y=x$ meets $y=0$ at $(0,0)$.
    * $y=2-x$ meets $y=0$ at $(2,0)$.

```{r}
# --- 3. Triangle (Two Lines) ---
tri2_data <- data.frame(x = c(0, 2, 1), y = c(0, 0, 1))
ggplot(tri2_data, aes(x, y)) +
  geom_polygon(fill="orange", color="black", alpha=0.5) +
  geom_segment(aes(x=0, y=0, xend=2.5, yend=2.5), linetype="dashed", color="gray") +
  geom_segment(aes(x=-0.5, y=2.5, xend=2.5, yend=-0.5), linetype="dashed", color="gray") +
  xlim(-0.5, 3) + ylim(0, 2.5) +
  labs(title = "Two Lines + Axis", x = "x", y = "y") +
  theme_minimal()
```


---

### Circular Regions
Inequality: $(x - h)^2 + (y - k)^2 \le r^2$

**How to Sketch:**

1.  Identify Center: The center is $(h, k)$.
2.  Identify Radius: The radius is $r$ (take the square root of the right hand side of the inequality).
3.  Test Point: Just like lines, you can test $(h,k)$. If the center satisfies the inequality (it usually does for $\le$), shade inside.

Example:
Sketch the region $x^2 + y^2 \le 4$.

* Center: $(0,0)$.
* Radius: $\sqrt{4} = 2$.
* Test Point: Is $0^2 + 0^2 \le 4$? Yes, so shade inside.

```{r}
# --- 4. Circle ---
theta <- seq(0, 2*pi, length.out=100)
circle_data <- data.frame(x = 2*cos(theta), y = 2*sin(theta))
ggplot(circle_data, aes(x, y)) +
  geom_polygon(fill="pink", color="black", alpha=0.5) +
  coord_fixed() + 
  labs(title = "Circular Region (r=2)", x = "x", y = "y") +
  theme_minimal()
```

---

### Practice

Now try to make some plots yourself! Sketch the following regions on an $x$-$y$ plane:

1.   Sketch the region defined by $0 \le x \le 2$ and $2 \le y \le 4$.

2.   Sketch the region defined by $x + y \le 3$, restricted to the first quadrant ($x \ge 0, y \ge 0$). Label the vertices.

3.  Sketch the region defined by $2x + y \le 4$, restricted to the first quadrant ($x \ge 0, y \ge 0$). Label the vertices.

3.   Sketch the region bounded by $y = x$, $y = 6 - 2x$, and the x-axis ($y = 0$).

4.   Sketch the region bounded by $y = 2x$, $y = 8 - 2x$, and the x-axis ($y = 0$). 

4.   Sketch the region defined by $x^2 + y^2 \le 9$.

5.   Sketch the region defined by $(x - 1)^2 + y^2 \le 1$.








