---
title: "Differentiation Basics"
author: "Jibo Shen"
output: 
  github_document:
    toc: false
---

```{r setup, include=FALSE}
# This chunk sets up the environment.
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.align="center")

library(ggplot2)
library(gridExtra)

# Common plotting theme for consistency
common_theme <- theme_minimal() + 
                theme(axis.title = element_text(face="bold"),
                      plot.title = element_text(face="bold", hjust = 0.5))

```

In this section, we review some differentiation basics. Differentiation is a fundamental tool in calculus. It is used to determine the instantaneous rate of change of a function. To put it simple, by inspecting the differentiation of a function, we can have a sense of how fast the function changes.

There are several notation systems for differentiation. In this course, we often use the following notation, which explicitly treats differentiation as an action applied to a function.

$$ \frac{d}{dx} \Big[ f(x) \Big] $$

You can read this as: "The derivative **with respect to x** ... of $f(x)$."

Note: The term "with respect to x" emphasizes the essential fact that $x$ is the variable changing, while all other letters are treated as constants.

You can also read in short the notation as: "d - d x ... of $f(x)$."

Another convenient notation is to write the differentiation of $f(x)$ as $f'(x)$.

---

## Part 1: Derivatives of Simple Functions

Here are the derivative of some simple function:

### The Constant Function
If $f(x) = c$, then the derivative is zero.

$$
\frac{d}{dx} c = 0
$$

### The Power function
For any real number $n$, the power function $f(x)=x^n$ has derivative:

$$
\frac{d}{dx} x^n = n x^{n-1}
$$

e.g.) $\frac{d}{dx} x^3 = 3x^{(3-1)}=3x^2$, $\frac{d}{dx} \frac{1}{x} = \frac{d}{dx} x^{-1} = (-1)x^{-1-1} = -\frac{1}{x^2}$

Note: Even though a power function has different behaviors depending on the value of $n$, the formula of derivative is the same.

### The Exponential Function
The Natural Exponential $f(x)=e^x$:
The natural exponential function is very unique because it is its own derivative.

$$
\frac{d}{dx} e^x = e^x
$$

The General Exponential $f(x)=a^x$:
For any base $a > 0$ ($a \neq 1$):

$$
\frac{d}{dx} a^x = a^x \log(a)
$$

### The Logarithmic Function
The derivative of the natural logarithm $f(x)=\log x$ is the reciprocal function.

$$
\frac{d}{dx} \log(x) = \frac{1}{x}
$$

For the general logarithm $f(x)=\frac{d}{dx} [\log_b(x)]$,

$$
\frac{d}{dx} [\log_b(x)] = \frac{1}{x \log(b)}
$$

### Trigonometric Functions
The derivatives of sine function $f(x)=\sin x$ and cosine function $f(x)=\cos x$ are cyclical.

$$
\frac{d}{dx} \sin(x) = \cos(x)
$$
$$
\frac{d}{dx} \cos(x) = -\sin(x)
$$

Note: Don't forget the negative sign when differentiating cosine!

## Part 2: Basic Properties

Differentiation is a linear operator, meaning it follows two predictable rules when dealing with sums and constants.

### Function multiplied by a constant
If you multiply a function by a constant $c$, the derivative is just multiplied by that same constant.

$$
\frac{d}{dx}[c \cdot f(x)] = c \cdot \frac{d}{dx}f(x)
$$

### Sum of two functions
The derivative of a sum is the sum of the derivatives.

$$
\frac{d}{dx}[f(x) + g(x)] = \frac{d}{dx}f(x) + \frac{d}{dx}g(x)
$$

e.g.) To differentiate $f(x) = x^2 + e^x$, you differentiate $x^2$ and $e^x$ separately, then add the results. You can also see that $\frac{d}{dx}[f(x) + c]=\frac{d}{dx}f(x)$.

Note: When doing differentiation, it is important to always keep in mind what is the changing variable, and what are the constants.

## Visualization of Derivatives

Visualizing a function alongside its derivative helps build intuition. Below, we look at the Natural Log function $f(x)=\log x$ (Red) and its derivative, $f(x)=1/x$ (Blue).

Notice that as the slope of the Red curve gets flatter (as $x$ increases), the value of the Blue curve gets closer to $0$.

```{r deriv_vis, fig.height=4, fig.width=6}
# Define the range for x (must be > 0 for log)
x_vals <- seq(0.1, 5, length.out = 200)

# Create a data frame with both f(x) and f'(x)
df_deriv <- data.frame(
  x = x_vals,
  Function = log(x_vals),       # f(x) = ln(x)
  Derivative = 1/x_vals         # f'(x) = 1/x
)

# Reshape for easier plotting with ggplot legend
library(tidyr)
df_long <- pivot_longer(df_deriv, cols = c("Function", "Derivative"), names_to = "Type", values_to = "Value")

ggplot(df_long, aes(x = x, y = Value, color = Type)) +
  geom_line(size = 1.2) +
  geom_hline(yintercept = 0, linetype="dashed", color = "gray") +
  labs(title = "Function vs. Derivative", 
       subtitle = "f(x) = log(x)  vs.  f'(x) = 1/x",
       y = "Value") +
  scale_color_manual(values = c("Derivative" = "#2980b9", "Function" = "#c0392b")) +
  common_theme
```

Let's look at another function $f(x) = x^3 - 3x$. In the plot, the function is colored by the slope (derivative) at each point. Note that the derivative is positive when $x<-1$ or $x>1$, is negative when $-1< x < 1$, and is zero at the point $x=-1$ and $x=1$.

```{r}
# Define the function and its derivative
f <- function(x) { x^3 - 3*x }
f_prime <- function(x) { 3*x^2 - 3 }

# Create data for plotting
x_vals <- seq(-2.5, 2.5, length.out = 100)
df_poly <- data.frame(
  x = x_vals,
  y = f(x_vals),
  slope = f_prime(x_vals) # We can use slope to color the curve
)

# Plot
ggplot(df_poly, aes(x, y)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray") +
  
  # Main function curve
  geom_line(size = 1.2, aes(color = slope)) +
  
  # Highlight the turning points (where derivative is zero)
  geom_point(data = data.frame(x = c(-1, 1), y = c(2, -2)), 
             size = 4, color = "black") +
  
  # Add tangent lines at the turning points
  geom_segment(aes(x = -1.5, y = 2, xend = -0.5, yend = 2), 
               color = "black", linetype = "dotted") +
  geom_segment(aes(x = 0.5, y = -2, xend = 1.5, yend = -2), 
               color = "black", linetype = "dotted") +
  
  scale_color_gradient2(low = "red", mid = "gray", high = "blue", 
                        midpoint = 0, name = "Slope value") +
  labs(title = "f(x) = x^3 - 3x", 
       subtitle = "Color indicates the slope (derivative)",
       y = "f(x)") +
  common_theme
```

