---
title: 'Activity: Visualizing Functions and Their Derivatives'
author: "Jibo Shen"
output: 
  github_document:
    toc: false
---

```{r setup, include=FALSE}
# This chunk sets up the environment.
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.align="center")

library(ggplot2)
library(gridExtra)

# Common plotting theme for consistency
common_theme <- theme_minimal() + 
                theme(axis.title = element_text(face="bold"),
                      plot.title = element_text(face="bold", hjust = 0.5))

```

While it is good to memorize the plot of some most common functions, like those mentioned in "Reading: Function Basics and Common Functions", sometimes we may encounter a function for the first time, and we need to sketch it from scratch.

## General Strategy

When you need to sketch a function $f(x)$ from scratch, it helps to follow a systematic checklist. By combining basic algebra with derivatives, you can map out any curve.

1. **Identify the Domain**
* Look for any $x$-values where the function cannot exist. 
* Watch out for: Division by zero (which creates vertical asymptotes or holes) and negative numbers under even roots.

2. **Find the Intercepts**
* $y$-intercept: Set $x = 0$ and solve for $y$. This is where the graph crosses the vertical axis. Note that there is at most one $y$-intercept.
* $x$-intercepts: Set $f(x) = 0$ and solve for $x$. These are the roots of the function, where it crosses the horizontal axis.

3. **Check End Behavior and Asymptotes**
* Horizontal Asymptotes: What happens to $y$ as $x$ gets extremely large (approaches $+\infty$) or extremely small (approaches $-\infty$)?
* Vertical Asymptotes: Draw dashed vertical lines at the $x$-values that make the denominator zero. The graph will curve sharply upward or downward as it gets close to these lines.

4. **Find Critical Points using $f'(x)$**
* Calculate the first derivative, $f'(x)$. Set it equal to $0$ and solve for $x$. Also find where $f'(x)$ is undefined.
* These $x$-values are your "critical points." They represent the exact locations where the graph temporarily flattens out—meaning they are the locations of your local maximums and minimums.

5. **Determine Increasing and Decreasing Intervals**
* Plot your critical points on a simple number line. Pick a "test number" in each section between the critical points and plug it into the derivative $f'(x)$.
* The Rule: 
    * If $f'(\text{test}) > 0$, the original function $f(x)$ is increasing (going up) in that interval.
    * If $f'(\text{test}) < 0$, the original function $f(x)$ is decreasing (going down) in that interval.

6. **Plot Key Points and Connect**
* Plot your $x$ and $y$ intercepts.
* Plug your critical $x$-values back into the original function $f(x)$ to find their exact $y$-coordinates, and plot those maximums and minimums.
* Draw a smooth curve connecting your plotted points, making sure the curve goes up where $f'(x)$ is positive, goes down where $f'(x)$ is negative, and respects any asymptotes.

---

### Example 1: Sketching a Polynomial Function

Let's sketch the function $f(x) = x^3 - 3x^2$.

1. Identify the Domain
* Since this is a polynomial, there are no denominators or square roots. The domain is all real numbers.

2. Find the Intercepts
* $y$-intercept: Plug in $x = 0$. We get $f(0) = 0^3 - 3(0)^2 = 0$. The graph crosses at $(0,0)$.
* $x$-intercepts: Set $f(x) = 0$. We get $x^3 - 3x^2 = 0$. Factoring out $x^2$ gives $x^2(x - 3) = 0$. This means $x = 0$ and $x = 3$. The graph crosses the horizontal axis at $(0,0)$ and $(3,0)$.

3. Check End Behavior and Asymptotes
* There are no asymptotes for polynomials. 
* As $x$ gets very large, $x^3$ dominates and the graph goes way up. As $x$ gets very negative, $x^3$ goes way down.

4. Find Critical Points using $f'(x)$
* The derivative is $f'(x) = 3x^2 - 6x$.
* Set it equal to $0$ to find critical points: $3x^2 - 6x = 0$, which factors to $3x(x - 2) = 0$.
* Our critical points are at $x = 0$ and $x = 2$.

5. Determine Increasing and Decreasing Intervals
* We check the intervals around $x = 0$ and $x = 2$.
* Test a number less than $0$ (like $x = -1$): $f'(-1) = 3(-1)^2 - 6(-1) = 9$. Positive, so the function is increasing.
* Test a number between $0$ and $2$ (like $x = 1$): $f'(1) = 3(1)^2 - 6(1) = -3$. Negative, so the function is decreasing.
* Test a number greater than $2$ (like $x = 3$): $f'(3) = 3(3)^2 - 6(3) = 9$. Positive, so the function is increasing.
* This tells us $x = 0$ is a local peak, and $x = 2$ is a local valley.

6. Plot Key Points and Connect
* We have our intercepts $(0,0)$ and $(3,0)$. 
* Let's find the exact coordinate of our valley at $x = 2$. Plug it into the original function: $f(2) = 2^3 - 3(2)^2 = 8 - 12 = -4$. So the valley is at $(2, -4)$.
* Now, connect the dots smoothly, rising to $(0,0)$, falling to $(2, -4)$, and rising through $(3,0)$.

```{r}
library(ggplot2)

df_ex1 <- data.frame(x = seq(-1.5, 4, length.out = 100))
df_ex1$y <- df_ex1$x^3 - 3 * df_ex1$x^2

ggplot(df_ex1, aes(x, y)) +
  geom_line(color = "blue", size = 1) +
  geom_hline(yintercept = 0, color = "black", size = 0.5) +
  geom_vline(xintercept = 0, color = "black", size = 0.5) +
  geom_point(data = data.frame(x = c(0, 3, 2), y = c(0, 0, -4)), 
             aes(x, y), color = "red", size = 3) +
  annotate("text", x = 0.5, y = 1, label = "Local Max (0,0)") +
  annotate("text", x = 2.5, y = -4.5, label = "Local Min (2,-4)") +
  labs(title = "Sketch of f(x) = x^3 - 3x^2", x = "x", y = "f(x)") +
  theme_minimal()
```

### Example 2: Sketching a Rational Function

Let's sketch the function $f(x) = \frac{x}{x - 2}$.

1. Identify the Domain
* The denominator cannot be zero, so $x - 2 \neq 0$, which means $x \neq 2$.

2. Find the Intercepts
* $y$-intercept: Plug in $x = 0$. We get $f(0) = \frac{0}{0 - 2} = 0$. The graph crosses at $(0,0)$.
* $x$-intercepts: Set $f(x) = 0$. A fraction is zero only when its numerator is zero. So, $x = 0$. 

3. Check End Behavior and Asymptotes
* Vertical Asymptote: Because the domain excludes $x = 2$, we draw a dashed vertical line at $x = 2$.
* Horizontal Asymptote: Look at the highest powers of $x$. They match (both are $x^1$). The ratio of their coefficients is $1/1 = 1$. So, as $x$ gets massive, the graph flattens out at the horizontal line $y = 1$.

4. Find Critical Points using $f'(x)$
* Using the quotient rule, the derivative is $f'(x) = \frac{1(x - 2) - x(1)}{(x - 2)^2} = \frac{-2}{(x - 2)^2}$.
* Set it equal to $0$: $\frac{-2}{(x - 2)^2} = 0$. This has no solution because the numerator is $-2$.
* There are no critical points. The graph has no standard peaks or valleys.

5. Determine Increasing and Decreasing Intervals
* We only need to test around our vertical asymptote at $x = 2$.
* Test a number less than $2$ (like $x = 0$): $f'(0) = \frac{-2}{(0 - 2)^2} = -0.5$. Negative, so it is decreasing.
* Test a number greater than $2$ (like $x = 3$): $f'(3) = \frac{-2}{(3 - 2)^2} = -2$. Negative, so it is also decreasing.
* The function is constantly decreasing on both sides of the asymptote.

6. Plot Key Points and Connect
* Draw the dashed asymptotes at $x = 2$ and $y = 1$.
* Plot the intercept at $(0,0)$.
* Since it decreases before $x = 2$ and passes through $(0,0)$, it must be trapped in the bottom-left section of the asymptotes.
* Since it decreases after $x = 2$, it must start from the top-right and curve down towards $y = 1$.

```{r}
library(ggplot2)

df_ex2_left <- data.frame(x = seq(-2, 1.9, length.out = 100))
df_ex2_right <- data.frame(x = seq(2.1, 6, length.out = 100))
df_ex2_left$y <- df_ex2_left$x / (df_ex2_left$x - 2)
df_ex2_right$y <- df_ex2_right$x / (df_ex2_right$x - 2)

ggplot() +
  geom_line(data = df_ex2_left, aes(x, y), color = "purple", size = 1) +
  geom_line(data = df_ex2_right, aes(x, y), color = "purple", size = 1) +
  geom_hline(yintercept = 0, color = "black", size = 0.5) +
  geom_vline(xintercept = 0, color = "black", size = 0.5) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray50") +
  geom_vline(xintercept = 2, linetype = "dashed", color = "gray50") +
  geom_point(aes(x = 0, y = 0), color = "red", size = 3) +
  coord_cartesian(ylim = c(-4, 6)) +
  labs(title = "Sketch of f(x) = x / (x - 2)", x = "x", y = "f(x)") +
  theme_minimal()
```

## Visualizing the Derivative
### The Tangent Line

When we say the derivative $f'(x)$ gives us the "slope" of the curve, what we are really talking about is the slope of the **tangent line** at any specific point.

A tangent line is a straight line that just barely grazes the curve of a function at one exact point, without crossing through it at that immediate location. It represents the exact direction the curve is heading at that very second.

If you pick a specific $x$-value, let's call it $x = a$:

1. The height of the curve is $f(a)$. The point on the graph is $(a, f(a))$.
2. The slope of the tangent line at that exact point is $f'(a)$.

**Finding the Tangent Line Equation:**

Once you have the point and the slope, you can find the full equation of the tangent line using the point-slope form from algebra:
$y - y_1 = m(x - x_1)$. Substituting our calculus terms, the equation of the tangent line is: $y - f(a) = f'(a)(x - a)$

Example:
Let's find and graph the tangent line to the function $f(x) = x^2$ at the point where $x = 1$.

1. Find the point: 
Plug $x = 1$ into the original function. $f(1) = 1^2 = 1$. Our point is $(1, 1)$.
2. Find the slope: 
Take the derivative. $f'(x) = 2x$. 
Plug $x = 1$ into the derivative to get the specific slope at that point. $f'(1) = 2(1) = 2$.
3. Write the equation:
Using our formula: $y - 1 = 2(x - 1)$. 
Simplifying this gives $y = 2x - 1$.

If you graph $f(x) = x^2$ and the line $y = 2x - 1$, you will see the line perfectly grazes the curve at the point $(1, 1)$.

```{r}
library(ggplot2)

df_tangent <- data.frame(x = seq(-2, 3, length.out = 100))
df_tangent$f_x <- df_tangent$x^2
df_tangent$tangent <- 2 * df_tangent$x - 1

ggplot(df_tangent, aes(x)) +
  geom_line(aes(y = f_x, color = "Function: f(x) = x^2"), size = 1.2) +
  geom_line(aes(y = tangent, color = "Tangent: y = 2x - 1"), size = 1, linetype = "dashed") +
  geom_point(aes(x = 1, y = 1), color = "black", size = 3) +
  geom_hline(yintercept = 0, color = "black", size = 0.5) +
  geom_vline(xintercept = 0, color = "black", size = 0.5) +
  coord_cartesian(ylim = c(-2, 5)) +
  scale_color_manual(name = "", values = c("Function: f(x) = x^2" = "#2980b9", "Tangent: y = 2x - 1" = "#e74c3c")) +
  annotate("text", x = 1.8, y = 1, label = "Point of Tangency (1, 1)") +
  labs(title = "Visualizing the Tangent Line", x = "x", y = "y") +
  theme_minimal() + theme(legend.position = "bottom")
```

### Visualizing the Changing Derivative

To see how the derivative $f'(x)$ represents the changing slope of the function, we can draw tangent lines at several different points along the curves from our previous examples.

Notice how the steepness and direction of each dashed tangent line perfectly matches the curve at that specific point.

**Tangent Lines for Example 1: $f(x) = x^3 - 3x^2$**

We will look at the tangents at $x = 0$ (a peak), $x = 1$ (a decreasing section), and $x = 3$ (a steep increasing section).

```{r}
library(ggplot2)

df_ex1 <- data.frame(x = seq(-1.5, 4, length.out = 100))
df_ex1$y <- df_ex1$x^3 - 3 * df_ex1$x^2

ggplot(df_ex1, aes(x, y)) +
  geom_line(color = "blue", size = 1) +
  geom_hline(yintercept = 0, color = "black", size = 0.5) +
  geom_vline(xintercept = 0, color = "black", size = 0.5) +
  
  # Tangent at x = 0: y = 0
  geom_abline(intercept = 0, slope = 0, linetype = "dashed", color = "red") +
  geom_point(aes(x = 0, y = 0), color = "red", size = 3) +
  
  # Tangent at x = 1: y = -3x + 1
  geom_abline(intercept = 1, slope = -3, linetype = "dashed", color = "darkgreen") +
  geom_point(aes(x = 1, y = -2), color = "darkgreen", size = 3) +
  
  # Tangent at x = 3: y = 9x - 27
  geom_abline(intercept = -27, slope = 9, linetype = "dashed", color = "orange") +
  geom_point(aes(x = 3, y = 0), color = "orange", size = 3) +
  
  coord_cartesian(ylim = c(-6, 6), xlim = c(-1, 4)) +
  labs(title = "Tangent Lines on f(x) = x^3 - 3x^2", x = "x", y = "f(x)") +
  theme_minimal()
```

**Tangent Lines for Example 2: $f(x) = \frac{x}{x - 2}$**

We will look at the tangents at $x = 0$, $x = 1$, and $x = 3$. Notice that because the derivative is always negative, all three tangent lines point downwards.

```{r}
library(ggplot2)

df_ex2_left <- data.frame(x = seq(-2, 1.9, length.out = 100))
df_ex2_right <- data.frame(x = seq(2.1, 6, length.out = 100))
df_ex2_left$y <- df_ex2_left$x / (df_ex2_left$x - 2)
df_ex2_right$y <- df_ex2_right$x / (df_ex2_right$x - 2)

ggplot() +
  geom_line(data = df_ex2_left, aes(x, y), color = "purple", size = 1) +
  geom_line(data = df_ex2_right, aes(x, y), color = "purple", size = 1) +
  geom_hline(yintercept = 0, color = "black", size = 0.5) +
  geom_vline(xintercept = 0, color = "black", size = 0.5) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray50") +
  geom_vline(xintercept = 2, linetype = "dashed", color = "gray50") +
  
  # Tangent at x = 0: y = -0.5x
  geom_abline(intercept = 0, slope = -0.5, linetype = "dashed", color = "red") +
  geom_point(aes(x = 0, y = 0), color = "red", size = 3) +
  
  # Tangent at x = 1: y = -2x + 1
  geom_abline(intercept = 1, slope = -2, linetype = "dashed", color = "darkgreen") +
  geom_point(aes(x = 1, y = -1), color = "darkgreen", size = 3) +
  
  # Tangent at x = 3: y = -2x + 9
  geom_abline(intercept = 9, slope = -2, linetype = "dashed", color = "orange") +
  geom_point(aes(x = 3, y = 3), color = "orange", size = 3) +
  
  coord_cartesian(ylim = c(-4, 6), xlim = c(-2, 5)) +
  labs(title = "Tangent Lines on f(x) = x / (x - 2)", x = "x", y = "f(x)") +
  theme_minimal()
```

---

## Practice

For each function below:
  1. Find the derivative $f'(x)$.
  2. Identify the critical points by setting $f'(x) = 0$.
  3. Sketch the general shape of the curve $f(x)$.
  4. Draw the dashed tangent line at the specified $x$-values.


1. $f(x) = e^{-x^2/2}$
* Draw tangents at: $x = -1$, $x = 0$, and $x = 1$.


2. Let $f(x) = x e^{-x}$, restricted to $x \ge 0$
* Draw tangents at: $x = 0$, $x = 1$, and $x = 2$.
* *Hint: Use the product rule for the derivative. To find the peak, set $f'(x) = 0$ and factor out the $e^{-x}$ term.*

3. Let $f(x) = x^2 - x^3$, restricted to the interval $0 \le x \le 1$
* Draw tangents at: $x = 0$, $x = \frac{2}{3}$, and $x = 1$.
* *Hint: This curve is completely trapped between $x=0$ and $x=1$. Notice how the tangent lines behave at the absolute boundaries of this domain.*

4. Let $f(x) = \frac{1}{\sqrt{x(1-x)}}$, restricted to the interval $0 < x < 1$
* Draw tangents at: $x = \frac{1}{4}$, $x = \frac{1}{2}$, and $x = \frac{3}{4}$.
* *Hint: Notice that the function is undefined at $x=0$ and $x=1$, creating vertical asymptotes. Finding where $f'(x) = 0$ will reveal a deep valley instead of a peak.*